---
title: "Model Validation"
author: "Lin Yu"
date: "2025-01-27"
format: 
  html: 
    toc: true
    embed-resources: true
---

```{r}
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(DT)
library(here)
```
## Recap of tutorial 1

Understand `random variable` through the following simulation: 
```{r}
n_sample <- 1000
beta0 <-  0.5
beta1 <- 2
x <- rnorm(n_sample,mean = 5, sd = 10)

random_error <- rnorm(n_sample,mean = 0, sd = 5)

y <- beta0 + beta1*x + random_error

pop_dat <- data.frame(y = y,
           x = x)

pop_dat %>% 
  ggplot(aes(x = x, y = y))+
  geom_point()+
  theme_bw()
```

observed **standard deviation of the data**:

```{r}
pop_dat %>% 
  ggplot(aes(y))+
  geom_histogram()+
  theme_bw()

sd(pop_dat$y)
```


```{r}
sample_size <- 100
sample_dat <- pop_dat[sample(c(1:sample_size),replace = FALSE),]

sample_dat %>% datatable()

sample_dat %>% 
  ggplot(aes(x = x, y = y))+
  geom_point()+
  theme_bw()
```

the standard deviation of the estimator ($\hat\beta$):
```{r}
glm_m <- glm(y~x, data= sample_dat)

set.seed(1017)
## lapply parallel computing: more effecient than a for loop:)

coef_dat <- lapply(1:100, function(i){
  sample_dat <- pop_dat[sample(100,replace = FALSE),]
  glm_m <- glm(y~x, data= sample_dat)
  coef(glm_m)[2]
})

coef_dat %>% head()


set.seed(1017) 
beta_hat_dat <- lapply(1:100, function(i) {
  sample_dat <- pop_dat[sample(seq_len(nrow(pop_dat)), ## sample from 1: n_sample 
                               sample_size, ## sample size
                               replace = FALSE), ]
  glm_m <- glm(y ~ x, data = sample_dat)
  coef(glm_m)[2]
}) %>% unlist()

beta_hat_dat %>%
  data.frame() %>% 
  ggplot(aes(.)) +
  geom_histogram()
```

```{r}
beta_hat <- NULL
for(sample_size in c(20,100,200,500,1000)){
  beta_hat_tmp <- lapply(1:100, function(i) {
  sample_dat <- pop_dat[sample(seq_len(nrow(pop_dat)), ## sample from 1: n_sample 
                               sample_size, ## sample size
                               replace = FALSE), ]
  glm_m <- glm(y ~ x, data = sample_dat)
  coef(glm_m)[2]
}) %>% unlist() 
  beta_hat <- cbind(beta_hat,beta_hat_tmp )
}

beta_hat <- beta_hat %>% data.frame()

colnames(beta_hat) <- paste0("sample size = ",c(20,100,200,500,1000))

beta_hat %>% 
  datatable()

apply(beta_hat,2, sd)
```

viz the distribution of $\hat \beta$:
```{r}
library(tidyr)
plot_dat <- beta_hat %>% 
  pivot_longer(cols = everything(),
    values_to = "beta_hat",
    names_to = "sample size",
    names_prefix = "sample size ="
               ) 
  
 plot_dat <- plot_dat %>%  mutate(`sample size` = as.numeric(`sample size`))

plot_dat$`sample size` <- factor(plot_dat$`sample size`, 
                                 levels = c("20","100", "200", "500", "1000"), 
                                 labels = c("20","100", "200", "500", "1000"))


  plot_dat %>% 
    ggplot(aes(x = beta_hat,color = `sample size`, fill = `sample size`)) +
    geom_histogram()+
    facet_grid(cols = vars(`sample size`),scales = "free_y")+
    theme_bw()+
    theme(legend.position = "bottom")

```


## Tutorial 2: Model Validation

Use the hwy2.RData file available from the course page. After attaching the rms package and doing the usual datadist()/options() task        
1. Fit the following model

```r
fit <- ols(rate~rcs(trks,4)+rcs(sigs1,4)+type+hwy,data=hwy2,x=TRUE,y=TRUE)
```

2. run both the ordinary bootstrap validation and the .632 bootstrap validation on this model. compare the results.

***

```{r}
#| warning: false
#| message: false
rm(list = ls())
library(rms)
library(dplyr)
## change your working directory as necessary
load(here("data","hwy2.RData"))

str(hwy2)

```

:::{.callout-note collapse="true"}
# How do you code factor/nominal variables?

When working with factor/nominal/categorical variables in R, preprocessing is necessary to ensure the data is properly handled by statistical models and algorithms. Many R functions automatically perform dummy coding when a variable is specified as a factor, typically using the `as.factor()` function.

_Side note: Dummy coding (sometimes confused with one-hot encoding in CS) is just one way to code factor variables. It is popular because of its ease of interpretation._ 
:::


:::{.callout-note collapse="true"}
# Validate the model

A complete process of model validation usually consists of two parts: **internal** and **external** validation. Ideally (when you’re lucky 😄), you will have two independent datasets, such as EHR data from two hospitals. You will first _build_ or ( _train_) and _validate_ your model internally using data from hospital A (the train-validation dataset). Then the model can be sent to hospital B for external validation (test set). If you do not have two datasets, you can manually divide the data into two parts, with one part mimicking the external dataset.       

- Split-sample or split-sample averaged(SSA)

- K-fold Cross-validation   

- Bootstrap: ordinary, .632

```{r}
# sample size 100k
set.seed(1017)
n_sample <- 100000
patient_id <- c(1:n_sample)

# sample with replacement
sample_dat <- sample(patient_id,size=n_sample,replace = TRUE) 
# proportion of samples selected in bootstrap sampling with replacement
((sample_dat%>% unique() %>% length())/n_sample )%>% round(3)

```

:::
```{r}
h.dd <- datadist(hwy2)
options(datadist="h.dd")
fit <- ols(rate~rcs(trks,4)+
             rcs(sigs1,4)+type+hwy,
           data=hwy2,x=TRUE,y=TRUE)
fit$x %>% head()
```


```{r}
set.seed(1017)
validate(fit, B=100)
```


```{r}
set.seed(1017)
validate(fit,method = ".632",B=100)
```

Bootstrapping with replacement leads to some data points being repeated in the bootstrap sample. As a result, the model may end up 'memorizing' these repeated points, which can cause it to perform better.
