---
title: "Survival Analysis 1"
author: "Lin Yu"
date: "2025-03-03"
format: 
  html: 
    toc: true
    embed-resources: true
---

```{r}
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(DT)
library(here)
```

## Recap of Tutorial 5

One question left from last week:

Suppose you fit a generalized linear model with X as covariate. You are asked to calculate the effect size of X on the outcome Y comparing X=3 to X=2, and the confidence interval of the effect size:

$$
g[E(Y\mid X) ] = \beta_1 + \beta_2 \mathbf{1}{\{X=2\}}+\beta_3 \mathbf{1}{\{X=3\}}
$$
```{r}
## More about confidence interval 
set.seed(1017)
x <- sample(1:3, size=1000, replace=TRUE, prob=c(0.2, 0.2, 0.6))
head(x)

dat_x <- data.frame(x = x %>% as.factor())

design_matrix <- model.matrix(~., data = dat_x)

head(design_matrix)

true_coef <- c(0.5,1,2)
true_Y_linear <-   design_matrix %*% true_coef

 random_error <- rnorm(  length(true_Y_linear), mean=0,sd=1) 
  
true_Y_cont <- true_Y_linear + random_error
  
   
  true_Y_prob <- exp(true_Y_linear)/(1+exp(true_Y_linear)) ## location param for logistic model
  true_Y_bin <- apply(true_Y_prob, 1, function(x) rbinom(n =1, size =1, prob = x)) ## for logistic model, sampling Y
  
  
sim_dat <-   data.frame(x = as.factor(x),
             true_Y_cont = true_Y_cont,
             true_Y_bin = true_Y_bin
             )

outcome_m <- glm(true_Y_bin ~ x, family = binomial("logit"), data = sim_dat)
summary(outcome_m)


```
### Method 1

Change the reference level:

```{r}
sim_dat$x <- relevel(sim_dat$x, ref = "2")

outcome_m2 <- glm(true_Y_bin ~ x, family = binomial("logit"), data = sim_dat)

summary(outcome_m2)

```

### Method 2

Ask yourself the following questions:

- what is the estimand/goal?

$$\beta_3 - \beta_2$$

- what is the estimator?

$$\hat\beta_3 - \hat \beta_2$$

- what is the distribution of MLE estimator?

$\hat\beta_3$ and $\hat \beta_2$ follow normal distribution.  (one of the nice properties of MLE estimators)

then we can easily calculate the variance of $$\hat\beta_3 - \hat \beta_2$$ using some math.

```{r}
cov_matrix <- vcov(outcome_m)

var2 <- cov_matrix[2,2]
var3 <- cov_matrix[3,3]
cov_b23 <- cov_matrix[2,3]
## refer to tutorial 4 for the calculation of var(A+B)
var_b3_minus_b2 <- var2 + var3  - 2*cov_b23

var_b3_minus_b2 %>% sqrt()

se <- var_b3_minus_b2 %>% sqrt()

se
```

the point estiamte is calculated as:
```{r}
point_est <- coef(outcome_m)[3]-coef(outcome_m)[2]
```

so we conclude that $\hat\beta_3 - \hat \beta_2$ follow a normal distribution with mean `r point_est` and se `r se`

Finaly question, what is the relationship between $\hat\beta_3 - \hat \beta_2$ and OR??

$$log(OR) = \hat\beta_3 - \hat \beta_2$$
You can calculate the point estimate and confidence interval for the OR as well! Hooray!
 
## Tutorial 6

In this and the following few tutorials, we will focus on survival analysis. 


```{r}
library(rms)
library(survival)
str(mgus2)
```
```{r}
fit1 <- npsurv(Surv(futime,death)~sex,data=mgus2)
library(dplyr)
tmp1 <- summary(fit1) %>% head(5) %>% unlist()
tmp <- summary(fit1)

km_dat <- data.frame(time = tmp$time,
                       risk = tmp$n.risk,
                     death = tmp$n.event) %>%
  mutate(surv_i = 1- death/risk)



survplot(fit1,n.risk=TRUE)

```


```{r}
print(fit1.mh <- survdiff(Surv(futime, death) ~ sex, data = mgus2))
```

```{r}
summary(fit1,times=12) # time is in months
```

```{r}
## risk difference at year 1
RD <- 0.9032- 0.851
RD + 1.96*sqrt(0.0118^2 +0.013^2)
RD - 1.96*sqrt(0.0118^2 +0.013^2)
```

we know that logHR follows a normal distribution. So the confidence interval for HR could be calculated as:

$$exp(logHR ± 1.96 \cdot se_{logHR}) = exp(logHR)\cdot exp(± 1.96 \cdot se_{logHR}) = HR \cdot exp(± 1.96 \cdot se_{logHR}) $$
```{r}
## HR
## HR <- (obs1/exp1)/(obs2/exp2)
HR <- (423/471)/(540/492)

HR

## variance_logHR <- (1/exp1) + (1/exp2)
se_logHR <- sqrt( (1/471) +(1/492) )


## lower bound
HR *exp(-1.96*se_logHR)



## 95% CI
## upper bound
HR *exp(1.96*se_logHR)


```
